Weighted Model Counting Without Parameter Variables

### 2021-03-05 doc/paper3/paper.tex
The core of the paper is in a decent shape

\paragraph{Related Work.}
Many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}. Some attempts have previously been made to
skip this intermediate stage of representing a Bayesian network as a WMC
instance and compile directly to a more convenient representation.
Specifically, direct compilation from Bayesian networks to SDDs
\cite{DBLP:conf/ecsqaru/ChoiKD13} and from structured Bayesian networks (i.e., a
generalisation of Bayesian networks) to probabilistic SDDs
\cite{shen2020modeling} have been considered. To the best of the authors'
knowledge, neither compilation approach has a publicly available implementation.

Our work is similar in that it introduces a new representation for computational
problems that is based on pseudo-Boolean functions---pseudo-Boolean projection
(PBP). We formally show that every WMC problem instance has a corresponding PBP
instance. We argue that this new format is more appropriate for WMC solvers
whose execution is build on pseudo-Boolean function manipulation, i.e.,
\textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC}
\cite{DBLP:conf/cp/DudekPV20}. Finally, we show how most methods of encoding
probabilistic inference on Bayesian networks to WMC can be adapted to the PBP
format. This transformation significantly improves inference speed,
outperforming all other considered algorithm-encoding combinations.

@@

\paragraph{Notes.}
\begin{itemize}
\item we also show correctness
\item and introduce a new notation
\end{itemize}

@@

% TODO: Explain two meanings of entailment in the definition of WMC.

Throughout the paper, we use set theoretic notation for many concepts in logic.
A \emph{clause} is a set of literals that are part of an implicit disjunction.
Similarly, a \emph{formula} in CNF is a set of clauses that are part of an
implicit conjunction. We identify a \emph{model} with a set of variables that
correspond to the positive literals in the model (and all other variables are
the negative literals of the model). We can then define a model's
\emph{cardinality} as the cardinality of this set. For example, let $\phi =
(\neg a \lor b) \land a$ be a propositional formula over variables $a$ and $b$.
Then an equivalent set-theoretic representation of $\phi$ is $\{ \{ \neg a, b
\}, \{ a \} \}$. Any subset of $\{ a, b \}$ is an interpretation of $\phi$,
e.g., $\{ a, b \}$ is a model of $\phi$ (written $\{ a, b \} \models \phi$) of
cardinality two, while $\emptyset$ is an interpretation but not a model.

\begin{definition}[WMC]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

\begin{definition}
  Let $\phi$ be a propositional formula over a set of variables $X$. Then $Y
  \subseteq X$ is a \emph{minimum-cardinality model} of $\phi$ if $Y \models
  \phi$ and $|Y| \le |Z|$ for all $Z \models \phi$.
\end{definition}

\begin{definition}[Minimum-Cardinality WMC]
  A \emph{minimum-cardinality WMC} instance has same tuple as a WMC instance,
  but its answer is defined to be $\sum_{Y \models \phi\text{, }|Y| = k}
  \prod_{Y \models l} w(l)$ (where $k = \min_{Y \models \phi} |Y|$) if $\phi$ is
  satisfiable, and zero otherwise.
\end{definition}

Our definition of WMC is largely based on the standard definition
\cite{DBLP:journals/ai/ChaviraD08}, but explicitly partitions variables into
indicator and parameter variables. In practice, we identify this partition in
one of two ways. If an encoding is generated by \textsf{Ace}, then variable
types are explicitly identified in the LMAP file generated alongside the
encoding. Otherwise, we define a variable $x$ to be a parameter variable if
$w(x) = w(\neg x) = 1$.

\subsection{Bayesian Network Encodings (mostly done)}

@@

\section{Pseudo-Boolean Functions} % TODO: start here
% TODO: need more related literature
% TODO: look at the addmc/background_reading dir

\paragraph{References.} % extensions and applications
\begin{itemize}
\item ADDs \cite{DBLP:journals/fmsd/BaharFGHMPS97}
\item AADDs \cite{DBLP:conf/ijcai/SannerM05}
\item CUDD \cite{somenzi1998cudd}
\item compiling BNs using VE \cite{DBLP:conf/ijcai/ChaviraD07}
\item FOADDs \cite{DBLP:journals/ai/SannerB09}
\item stochastic planning \cite{DBLP:conf/uai/HoeySHB99}
\item XADDs \cite{DBLP:conf/uai/SannerDB11}
\item another connection between BNs and ADDs \cite{DBLP:conf/icml/ZhaoMP15}
\item and one more \cite{DBLP:conf/uai/GogateD11}
\item maybe: \cite{DBLP:conf/aaai/FargierMNS14}
\end{itemize}

\paragraph{Notation.} For any propositional formula $\phi$ over a set of
variables $X$ and $p, q \in \mathbb{R}$, let $[\phi]^p_q\colon 2^X \to
\mathbb{R}$ be a pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$.

\begin{definition}[Operations] \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions, $x, y \in X$,
  $Y = \{y_i\}_{i=1}^n \subseteq X$, and $r \in \mathbb{R}$. Operations such as
  addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (f \cdot g)(Y)
    \coloneqq f(Y) \cdot g(Y).
  \]
  Note that this means that binary operations on pseudo-Boolean functions
  inherit properties such as associativity and commutativity. By not
  distinguishing between a real number and a pseudo-Boolean function that always
  returns that number, we can use the same definitions to define \emph{scalar}
  operations as
  \[
    (r+f)(Y) = r+f(Y), \quad \text{and} \quad (r \cdot f)(Y) = r \cdot f(Y).
  \]

@@

Below we list some properties of the operations on pseudo-Boolean functions
discussed in this section that can be conveniently represented using our syntax.
The proofs of all these properties follow directly from the definitions.

@@

\section{Pseudo-Boolean Projection (mostly done)}

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X_I, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
%  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
%  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
%  replace every variable $v$ in $F$ with $\rename{$v$}$\;
%  \Return{$(F, X_I, \omega)$}\;
%  \Fn{\rename{$v$}}{
%    $S \gets \{u \in X_P \mid u \le v\}$\;
%    \lIf{$S = \emptyset$}{\Return{$v$}}
%    \Return{$v - d(\max S)$}\;
%  }
\end{algorithm}

@@

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$,
  \[
    (\exists_X f)(\emptyset) = \sum_{Y \subseteq X} f(Y).
  \]
\end{lemma}
\begin{proof}
  If $X = \{x\}$, then
  \[
    (\exists_xf)(\emptyset) = (f|_{x=1} + f|_{x=0})(\emptyset) =
    f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset) = \sum_{Y \subseteq \{x\}} f(Y).
  \]
  This easily extends to $|X| > 1$ by the definition of projection on sets of
  variables.
\end{proof}

\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

@@

\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{cond:d02,cond:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1,
  \]
  and indeed
  \[
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
      \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
    \right]_1^{w(p)}.
  \]
  This finishes the proof of the correctness of the first `foreach' loop under
  \cref{cond:d02}.

@@

  On the one hand, if $Z \subseteq X_I \cup X_P$ is a minimum-cardinality model
  of $\phi$, then
  \[
    \left(\prod_{f \in F}\right)(Z \cap X_I) \ne 0
  \]
  under the given assumptions. On the other hand, if $Y \subseteq X_I$ is such
  that
  \[
    \left(\prod_{f \in F}\right)(Y) \ne 0,
  \]
  then $Y \models \{ c \in \phi \mid c \cap X_P = \emptyset \}$. Let $Y
  \subseteq Z \subseteq X_I \cup X_P$ be the smallest superset of $Y$ such that
  $Z \models \phi$ (it exists by \cref{cond:2b3} of \cref{thm:correctness}). We
  need to show that $Z$ has minimum cardinality. Let $Y'$ and $Z'$ be defined
  equivalently to $Y$ and $Z$. We will show that $|Z| = |Z'|$. Note that $|Y| =
  |Y'|$ by \cref{cond:22}, and $|Z \setminus Y| = |Z' \setminus Y'|$ by
  \cref{cond:23}. Combining that with the general property that $|Z| = |Y| + |Z
  \setminus Y|$ finishes the first part of the proof.

@@

\section{Experimental Evaluation (mostly done)}
% TODO: what are the encodings compiled with?

@@

\textsf{DPMC} is run with tree decomposition based planning and algebraic
decision diagram based execution---the best-performing combination in the
original set of experiments \cite{DBLP:conf/cp/DudekPV20}. We use a single
iteration of \textsf{htd} \cite{DBLP:conf/cpaior/AbseherMW17} to generate
approximately optimal tree decompositions---we found that this configuration
is efficient enough to handle huge instances, and yet the width of the
returned decomposition is unlikely to differ from optimal by more than one or
two. We also enabled \textsf{DPMC}'s greedy mode. This mode (which was not part
of the original paper \cite{DBLP:conf/cp/DudekPV20}) optimises the order in
which pseudo-Boolean functions are multiplied by prioritising functions with
small representations.

>>>>

% NOTE: What is the problem? Why is it interesting and important?
WMC has emerged as a powerful computational framework for problems in many
domains, e.g., probabilistic graphical models such as Bayesian networks and
Markov networks
\cite{DBLP:conf/ecai/BartKLM16,DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06,DBLP:conf/kr/Darwiche02,DBLP:conf/aaai/SangBK05},
neuro-symbolic artificial intelligence \cite{DBLP:conf/icml/XuZFLB18},
probabilistic programs \cite{DBLP:journals/pacmpl/HoltzenBM20}, and
probabilistic logic programs \cite{DBLP:journals/tplp/FierensBRSGTJR15}. It has
been extended to support continuous variables \cite{DBLP:conf/ijcai/BellePB15},
infinite domains \cite{DBLP:conf/aaai/Belle17}, first-order logic
\cite{DBLP:journals/cacm/GogateD16,DBLP:conf/ijcai/BroeckTMDR11}, and arbitrary
semirings \cite{DBLP:journals/ijar/BelleR20,DBLP:journals/japll/KimmigBR17}.

% NOTE: What's wrong with previous proposed solutions? How does mine differ?
% TODO: Our work is similar to \cite{DBLP:conf/uai/GogateD10} who suggest using
% a collection of soft (i.e., with weights) and hard clauses
% TODO: how different are the semantics?
% But:
% 1) we can handle arbitrary formulas and not just clauses
% 2) we assign two weights, not one
% 3) the semantics of these weights are completely different, similar to the
% ones from Markov networks. It is thus difficult to reuse the same formulation
% for a different WMC algorithm.
% 4) Since we present a WMC->PBP algorithm, we don't need to invent new
% encodings for each problem and we can benefit from all the good ideas that
% went into designing WMC encodings

Many WMC inference algorithms such as
\textsf{Ace}\footnote{\url{http://reasoning.cs.ucla.edu/ace/}}, \textsf{c2d}
\cite{DBLP:conf/ecai/Darwiche04}, and \textsf{miniC2D}
\cite{DBLP:conf/ijcai/OztokD15} work by compilation to tractable representations
such as arithmetic circuits, deterministic, decomposable negation normal form
\cite{DBLP:journals/jancl/Darwiche01}, and sentential decision diagrams (SDDs)
\cite{DBLP:conf/ijcai/Darwiche11}. Some attempts have previously been made to
skip this intermediate stage of representing a Bayesian network as a WMC
instance and compile directly to a more convenient representation.
Specifically, direct compilation from Bayesian networks to SDDs
\cite{DBLP:conf/ecsqaru/ChoiKD13} and from structured Bayesian networks (i.e., a
generalisation of Bayesian networks) to probabilistic SDDs
\cite{shen2020modeling} have been considered. To the best of the authors'
knowledge, neither compilation approach has a publicly available implementation.

% NOTE: What are the key components of my approach and results? Also include any
% specific limitations. What about your project is new or different from
% existing solutions?
Our work is similar in that it introduces a new representation for computational
problems that is based on pseudo-Boolean functions---pseudo-Boolean projection
(PBP). We formally show that every WMC problem instance has a corresponding PBP
instance. We argue that this new format is more appropriate for WMC solvers
whose execution is build on pseudo-Boolean function manipulation, i.e.,
\textsf{ADDMC} \cite{DBLP:conf/aaai/DudekPV20} and \textsf{DPMC}
\cite{DBLP:conf/cp/DudekPV20}. Finally, we show how most methods of encoding
probabilistic inference on Bayesian networks to WMC can be adapted to the PBP
format. This transformation significantly improves inference speed,
outperforming all other considered algorithm-encoding combinations.

@@

% NOTE: List of refutable contributions (with forward references to the rest of
% the paper) What is the outcome of the project?
\paragraph{Notes.}
\begin{itemize}
\item we also show correctness
\item and introduce a new notation
\item TODO: briefly mention what a pseudo-Boolean function is
\end{itemize}

@@

We begin with an overview of some notation and terminology. Throughout the
paper, we use set theoretic notation for many concepts in logic. A \emph{clause}
is a set of literals that are part of an implicit disjunction. Similarly, a
\emph{formula} in CNF is a set of clauses that are part of an implicit
conjunction. We identify a \emph{model} with a set of variables that correspond
to the positive literals in the model (and all other variables are the negative
literals of the model). We can then define the \emph{cardinality} of a model as
the cardinality of this set. For example, let $\phi = (\neg a \lor b) \land a$
be a propositional formula over variables $a$ and $b$. Then an equivalent
set-theoretic representation of $\phi$ is $\{ \{ \neg a, b \}, \{ a \} \}$. Any
subset of $\{ a, b \}$ is an interpretation of $\phi$, e.g., $\{ a, b \}$ is a
model of $\phi$ (written $\{ a, b \} \models \phi$) of cardinality two, while
$\emptyset$ is an interpretation but not a model. We can now formally define
WMC.

\begin{definition}[WMC]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of \emph{indicator} variables, $X_P$ is the set of \emph{parameter}
  variables (with $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula
  in CNF over $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in
  X_I \cup X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of
  the instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

That is, the answer to a WMC instance is the sum of the weights of all models of
$\phi$, where the weight of a model is defined as the product of the weights of
all (positive and negative) literals in it. Our definition of WMC is largely
based on the standard definition \cite{DBLP:journals/ai/ChaviraD08}, but
explicitly partitions variables into indicator and parameter variables. In
practice, we identify this partition in one of two ways. If an encoding is
generated by \textsf{Ace}, then variable types are explicitly identified in the
LMAP file generated alongside the encoding. Otherwise, we define a variable $x$
to be a parameter variable if $w(x) = w(\neg x) = 1$. Next, we formally define a
variation of the WMC problem used by some of the Bayesian network encodings
\cite{DBLP:conf/ijcai/ChaviraD05,DBLP:conf/sat/ChaviraD06}.

\begin{definition}
  Let $\phi$ be a formula over a set of variables $X$. Then $Y \subseteq X$ is a
  \emph{minimum-cardinality model} of $\phi$ if $Y \models \phi$ and $|Y| \le
  |Z|$ for all $Z \models \phi$.
\end{definition}

\begin{definition}[Minimum-Cardinality WMC]
  A \emph{minimum-cardinality WMC} instance consists of the same tuple as a WMC
  instance, but its answer is defined to be $\sum_{Y \models \phi\text{, }|Y| =
    k} \prod_{Y \models l} w(l)$ (where $k = \min_{Y \models \phi} |Y|$) if
  $\phi$ is satisfiable, and zero otherwise.
\end{definition}

\subsection{Bayesian Network Encodings}

@@

\section{Pseudo-Boolean Functions}

In this work we propose a more expressive representation for WMC based on
pseudo-Boolean functions. Pseudo-Boolean functions, most commonly represented as
algebraic decision diagrams (ADDs) \cite{DBLP:journals/fmsd/BaharFGHMPS97}
(although a tensor-based approach has also been suggested
\cite{DBLP:journals/corr/abs-1908-04381,DBLP:conf/cp/DudekPV20}), have seen
extensive use in value iteration for Markov decision processes
\cite{DBLP:conf/uai/HoeySHB99}, both exact and approximate Bayesian network
inference \cite{DBLP:conf/ijcai/ChaviraD07,DBLP:conf/uai/GogateD11}, and
sum-product network \cite{DBLP:conf/uai/PoonD11} to Bayesian network conversion
\cite{DBLP:conf/icml/ZhaoMP15}. ADDs have been extended to compactly represent
additive and multiplicative structure \cite{DBLP:conf/ijcai/SannerM05},
sentences in first-order logic \cite{DBLP:journals/ai/SannerB09}, and continuous
variables \cite{DBLP:conf/uai/SannerDB11}, the last of which was also applied to
weighted model integration, i.e., the WMC extension for continuous variables
\cite{DBLP:conf/ijcai/BellePB15,DBLP:conf/ijcai/KolbMSBK18}.

Since two-valued pseudo-Boolean functions will be used extensively henceforth,
we introduce some notation. For any propositional formula $\phi$ over a set of
variables $X$ and $p, q \in \mathbb{R}$, let $[\phi]^p_q\colon 2^X \to
\mathbb{R}$ be a pseudo-Boolean function defined as
\[
  [\phi]^p_q(Y) \coloneqq
  \begin{cases}
    p & \text{if } Y \models \phi \\
    q & \text{otherwise}
  \end{cases}
\]
for any $Y \subseteq X$. Next, we define some useful operations on
pseudo-Boolean functions.
% TODO: mention where the operations are from
% TODO: should I consider different domains?
% TODO: can I say that a domain automatically shrinks (w.r.t. x) if f|x=0 =
% f|x=1?

\begin{definition}[Operations] \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions, $x, y \in X$,
  $Y = \{y_i\}_{i=1}^n \subseteq X$, and $r \in \mathbb{R}$. Operations such as
  addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (f \cdot g)(Y)
    \coloneqq f(Y) \cdot g(Y).
  \]
  Note that this means that binary operations on pseudo-Boolean functions
  inherit properties such as associativity and commutativity. By regarding a
  real number as a constant pseudo-Boolean function, we can reuse the same
  definitions to define \emph{scalar} operations as
  \[
    (r+f)(Y) = r+f(Y), \quad \text{and} \quad (r \cdot f)(Y) = r \cdot f(Y).
  \]

@@

Finally, below we list some properties of the operations on pseudo-Boolean
functions discussed in this section that can be conveniently represented using
our syntax. The proofs of all these properties follow directly from the
definitions.

@@

\section{Pseudo-Boolean Projection}

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X_I, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
%  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
%  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{$v \in X_I$ such that $\{[v]_1^p, [\neg v]_1^q\}
    \subseteq F$ for some $p$ and $q$ \label{line:foreach2start}}{
    $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
    \}$\; \label{line:foreach2end}
  }
%  replace every variable $v$ in $F$ with $\rename{$v$}$\;
%  \Return{$(F, X_I, \omega)$}\;
%  \Fn{\rename{$v$}}{
%    $S \gets \{u \in X_P \mid u \le v\}$\;
%    \lIf{$S = \emptyset$}{\Return{$v$}}
%    \Return{$v - d(\max S)$}\;
%  }
\end{algorithm}

@@

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$, we have that
  $(\exists_X f)(\emptyset) = \sum_{Y \subseteq X} f(Y)$.
\end{lemma}
\begin{proof}
  If $X = \{x\}$, then
  \[
    (\exists_xf)(\emptyset) = (f|_{x=1} + f|_{x=0})(\emptyset) =
    f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset) = \sum_{Y \subseteq \{x\}} f(Y).
  \]
  This easily extends to $|X| > 1$ by the definition of projection on sets of
  variables.
\end{proof}

\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  Let $f = \prod_{c \in \phi} [c]_0^1$, and $g = \prod_{x \in X_I \cup X_P}
  [x]_{w(\neg x)}^{w(x)}$. Then the answer to the WMC instance
  \eqref{eq:new_wmc} is
  \[
    (\exists_{X_I \cup X_P} fg)(\emptyset) = \sum_{Y \subseteq X_I \cup X_P}
    (fg)(Y) = \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y)
  \]
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that $\sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models
    \phi} \prod_{Y \models l} w(l)$ as required.
\end{proof}

@@

\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is $\left(\exists_{X_I \cup X_P} \left(\prod_{c
        \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg
      x)}^{w(x)} \right)(\emptyset)$. Since both \cref{cond:d02,cond:bklm16}
  ensure that each clause in $\phi$ has at most one parameter variable, we can
  partition $\phi$ into $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c)
  \cap X_P = \emptyset \}$ and $\phi_p \coloneqq \{ c \in \phi \mid
  \mathtt{Vars}(c) \cap X_P = \{ p \} \}$ for all $p \in X_P$. We can then use
  \cref{thm:early} to reorder the answer into $\left(\exists_{X_I} \left(
      \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right) \left( \prod_{c \in
        \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p [p]_{w(\neg
      p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset)$.

@@

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that $\left[ \bigwedge_{i = 1}^n l_i
  \right]_1^{w(p)} = \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow
    \bigwedge_{i=1}^n l_i \right]_0^1$, and indeed
  \[
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[
      \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i
    \right]_1^{w(p)}.
  \]
  This finishes the proof of the correctness of the first `foreach' loop under
  \cref{cond:d02}.

@@

  On the one hand, if $Z \subseteq X_I \cup X_P$ is a minimum-cardinality model
  of $\phi$, then $\left(\prod_{f \in F}\right)(Z \cap X_I) \ne 0$ under the
  given assumptions. On the other hand, if $Y \subseteq X_I$ is such that
  $\left(\prod_{f \in F}\right)(Y) \ne 0$, then $Y \models \{ c \in \phi \mid c
  \cap X_P = \emptyset \}$. Let $Y \subseteq Z \subseteq X_I \cup X_P$ be the
  smallest superset of $Y$ such that $Z \models \phi$ (it exists by
  \cref{cond:2b3} of \cref{thm:correctness}). We need to show that $Z$ has
  minimum cardinality. Let $Y'$ and $Z'$ be defined equivalently to $Y$ and $Z$.
  We will show that $|Z| = |Z'|$. Note that $|Y| = |Y'|$ by \cref{cond:22}, and
  $|Z \setminus Y| = |Z' \setminus Y'|$ by \cref{cond:23}. Combining that with
  the general property that $|Z| = |Y| + |Z \setminus Y|$ finishes the first
  part of the proof.

@@

\section{Experimental Evaluation}
% TODO: what are the encodings compiled with?

@@

\textsf{DPMC} is run with tree decomposition based planning and ADD based
execution---the best-performing combination in the original set of experiments
\cite{DBLP:conf/cp/DudekPV20}. We use a single iteration of \textsf{htd}
\cite{DBLP:conf/cpaior/AbseherMW17} to generate approximately optimal tree
decompositions---we found that this configuration is efficient enough to handle
huge instances, and yet the width of the returned decomposition is unlikely to
differ from optimal by more than one or two. We also enabled \textsf{DPMC}'s
greedy mode. This mode (which was not part of the original paper
\cite{DBLP:conf/cp/DudekPV20}) optimises the order in which pseudo-Boolean
functions are multiplied by prioritising functions with small representations.
